---
title: "Misa Kasparcova STAT 490 HW3"
author: "Misa Kasparcova"
date: "2025-10-09"
output: html_document
---

## Question 1:

Use fixedpoint to find the fixed point of cos x. Start with x0 = 0 as your
initial guess (the answer is 0.73908513).
Now use newtonraphson to find the root of cos x − x, starting with x0 = 0
as your initial guess. Is it faster than the fixed-point method?

Conduct fixedpoint algorithm.
```{r}
tol = 1e-12
maxit = 1000
x = 0
x_new = 0

for (i in 1:maxit) {
  x_new = cos(x)
  if (abs(x_new - x) < tol) {
    conv_iter = i
    x = x_new
  }
  x = x_new
}

x
```

Conduct newtonraphson algorithm.
```{r}
f = function(x) cos(x) - x
f_der = function(x) -sin(x) - 1

tol = 1e-12
maxit = 1000
x = 0
conv_iter = maxit

for (i in 1:maxit) {
  x_new = x - f(x)/f_der(x)
  if (abs(x_new - x) < tol) {
  }
  x = x_new
}

x
```

From conducting the code, we see that the newtonraphson method is faster.


## Question 2:

Write a program, using both newtonraphson.r and fixedpoint.r for guidance, 
to implement the secant root-finding method:

$$
x_{n+1} = x_n - f(x_n) \, \frac{x_n - x_{n-1}}{f(x_n) - f(x_{n-1})}.
$$

First test your program by finding the root of the function cos x − x. Next
see how the secant method performs in finding the root of log x − exp(−x)
using $x_0$ = 1 and $x_1$ = 2. Compare its performance with that of the other
two methods. Write a function secant_show.r that plots the sequence of iterates
produced by the secant algorithm.


##DO QUESTION
```{r}

```




## Question 3:

The mean value theorem states that for a function f which is continuous
on [a, b] and differentiable on (a, b), there exists a point $\theta \in$ (a, b)
such that

$$
f'(\theta) = \frac{f(b) - f(a)}{b - a}.
$$

Supposing that f, f′, a and b are given, explain how to use a root-finding
algorithm to find $\theta$.

```{r}

```


## Question 4:

Make your own R function to implement the secant algorithm. Your R function
should have a form:
mysecant <- function(f, x0, x1, tol = 1e-8){
expressions...
}

Test your code by solving logx − $e^{-x}$ + 1 = 0 with respect to x. In doing so, you should draw
a graph of f(x) = logx − $e^{-x}$ + 1 = 0 over x $\in$ [0, 1] and set up two initial values appropriately.

```{r}

```



## Question 5:

Let \( p \) be the quadratic \( p(x) = c_0 + c_1 x + c_2 x^2 \).
Simpson’s rule uses a quadratic to approximate a given function \( f \) over
two adjacent intervals, then uses the integral of the quadratic to
approximate the integral of the function.

(a) Show that

$$
\int_{-h}^{h} p(x) \, dx = 2h c_0 + \frac{2}{3} c_2 h^3.
$$

```{r}

```



(b) Write down three equations that constrain the quadratic to pass through
the points \( (-h, f(-h)) \), \( (0, f(0)) \), and \( (h, f(h)) \),
then solve them for \( c_0 \) and \( c_2 \).

```{r}

```



(c) Show that

$$
\int_{-h}^{h} p(x) \, dx = \frac{h}{3} \left[ f(-h) + 4f(0) + f(h) \right].
$$

```{r}

```


## Question 6:

The Riemann approximation to an integral $I = \int_a^b f(x) \, dx$ uses 
rectangles.  
For a partition of size $n$ we have the approximation  

$$
\hat{I} = \sum_{i=0}^{n-1} f(a + ih)h
$$  

where $h = (b - a)/n$ is the width of each interval in the partition.  

What is the difference between the Riemann and trapezoidal approximations 
(using the same partition)?  

It is known that if $-\infty < a < b < \infty$ and $f$ is continuous on 
$[a, b]$, then $\hat{I}$ converges as $n \to \infty$. Using this result, 
conclude that the trapezoidal approximation converges as $n \to \infty$, 
to the same value.




## Question 7:

Use the golden-section search algorithm to find all of the local maxima of
the function

$$
f(x) =
\begin{cases}
0, & x = 0 \\
|x| \log\left(\frac{|x|}{2}\right) e^{-|x|}, & \text{o/w}
\end{cases}
$$
within the interval [−10, 10].
Hint: plotting the function first will give you a good idea where to look.


