---
title: "Misa Kasparcova STAT 433 HW5"
author: "Misa Kasparcova"
date: "2025-11-05"
output:
  pdf_document: default
  html_document: default
---

## Question 1:

Let $f$ be a standard normal density. Use the accept-reject method to generate 
10000 samples from the standard normal distribution, $f$. In doing so, we 
select the standard Cauchy, $g(x)$, as the candidate density,

$$
g(x) = \frac{1}{\pi} \frac{1}{1 + x^2}, \quad x \in \mathbb{R}
$$

(a) Explain the implementation of the accept-reject method in detail. Plot a
histogram of the sample to verify it looks like a normal distribution. 
Generate 10000 random numbers.


The accept-reject method is used when the inverse transformation method fails 
to generate the required random variables. To utilize this method, we need to 
know the functional form of the density (f), but we utilize a simpler form of 
the density as (g), which is called the candidate density. This candidate 
density must have f and g as compatible supports and there must be a contant M 
such that f(x)/g(x) is less than or equal to M for all x. 


Function of normal distribution
```{r}
f = function(x) (1/sqrt(2*pi)) * exp(-x^2 / 2)
```

Function of standard Cauchy
```{r}
g = function(x) (1/pi) * (1 / (1 + x^2))
```

Obtain the max of f(x)/g(x)
```{r}
x_seq = seq(-10, 10, length.out = 10000)
ratio = f(x_seq) / g(x_seq)
c = max(ratio)
c
```

Generate 10000 random numbers in the sample
```{r}
set.seed(123)
n = 10000
samples = numeric(0)

while(length(samples) < n) {
  x = rcauchy(1)
  u = runif(1)
  if (u <= f(x) / (c * g(x))) {
    samples = c(samples, x)
  }
}
```


Plot the normal distribution and standard Cauchy
```{r}
hist(samples,  probability = TRUE,
     main = "Histogram of Sample", xlab = "x")
curve(dnorm(x), add = TRUE, lwd = 3)
```




(b) To generate 10000 random numbers in (a), how many are drawn from the
candidate density? Compare this number with its theoretical mean.

Obtain normal distribution and standard Cauchy
```{r}
set.seed(123)
f = function(x) dnorm(x)
g = function(x) dcauchy(x)
```

Obtain the max of f(x)/g(x)
```{r}
x = seq(-10, 10, length.out = 10000)
ratio = f(x) / g(x)
c = max(ratio)
c
```


```{r}
n = 10000
samples = numeric(0)
cand = 0

while (length(samples) < n) {
  x = rcauchy(1)          
  u = runif(1)            
  cand = cand + 1
  if (u <= f(x) / (c * g(x))) {
    samples = c(samples, x)
  }
}

cat("Candidate draws is", cand, "\n")
cat("Theoretical expected is", n * c, "\n")
```

We observe that the candidate draws is 15184, which is very close to the 
expected theoretical mean of 15203.46.



(c) Noting that $E(X) = 0$ and $var(X) = 1$, provide the sample mean and 
variance of the generated sample in (a). Comment on your result.

Obtain sample mean and variance from part (a)
```{r}
sample_mean = mean(samples)
sample_var  = var(samples)
cat("Sample mean is", sample_mean, "\n")
cat("Sample variance is", sample_var, "\n")
```

The simulated sample mean of 0.00116 is extremely close to the theoretical mean 
of 0, and the sample variance of 0.98646 is very close to the theoretical 
variance of 1. We can observe that the results confirm that the accept–reject 
method successfully generated a sample that is in line with the standard normal 
distribution.



## Question 2:

Consider the discrete random variable with pmf given by:

$$
P(X = 1) = 0.1, \quad P(X = 2) = 0.3, \quad P(X = 5) = 0.6
$$

Plot the cdf for this random variable. Write a program to simulate a random 
variable with this distribution, using the built-in function runif(1).

Obtain the pmf
```{r}
x = c(1, 2, 5)
p = c(0.1, 0.3, 0.6)
```

Obtain cdf
```{r}
cdf = F = c(p[1], p[1] + p[2], 1)
cdf
```

Plot the cdf
```{r}
plot(x, cdf, type = "s", lwd = 2,
     xlab = "x", ylab = "F(x)", main = "CDF of X")
points(x, cdf, pch = 15, col = "red")
```


Simulate a random variable using runif(1)
```{r}
rand_var = runif(1)
X = x[which.max(rand_var <= cdf)]
cat("X =", X, "\n")
```



## Question 3:

Consider the continuous random variable with pdf given by:

$$
f(x) = 3(x - 1)^2, \quad 1 < x < 2
$$

Plot the cdf for this random variable.
Show how to simulate a rv with this cdf using the inversion method.


Create pdf of function
```{r}
pdf = function(x) {
  3 * (x - 1)^2
}
```


Create cdf of function by obtaining integral from 1 to x
```{r}
cdf =  function(x) {
  3 * ((x - 1)^3) / 3  
}
```


Plot of cdf variable
```{r}
curve(cdf(x), from = 1, to = 2,
      main = "CDF of f(x)",
      xlab = "x", ylab = "F(x)")
```


Simulate a random variable with this cdf using the inversion method
```{r}
rand_var2 = runif(1)
X2 = 1 + rand_var2^(1/3)
cat("X =", X2, "\n")
```



