---
title: "Misa Kasparcova STAT 433 HW3"
author: "Misa Kasparcova"
date: "2025-10-09"
output:
  pdf_document: default
  html_document: default
---

## Question 1:

Use fixedpoint to find the fixed point of cos x. Start with x0 = 0 as your
initial guess (the answer is 0.73908513).
Now use newtonraphson to find the root of cos x - x, starting with x0 = 0
as your initial guess. Is it faster than the fixed point method?



Conduct fixedpoint algorithm.
```{r}
fixed_point = function(g, x0, tol = 1e-12, maxit = 1000) {
  xs = numeric(maxit + 1)
  xs[1] = x0
  x = x0
  conv_iter = NA_real_
  for (i in 1:maxit) {
    x_new = g(x)
    xs[i + 1] = x_new
    if (is.na(conv_iter) && abs(x_new - x) < tol) conv_iter = i
    x = x_new
  }
  if (is.na(conv_iter)) conv_iter = maxit
  list(root = xs[conv_iter + 1], iterations = conv_iter, values = xs[1:(conv_iter+1)])
}

g_cos = function(x) cos(x)
fp_res = fixed_point(g_cos, x0 = 0, tol = 1e-12, maxit = 1000)

signif(fp_res$root, 12)
fp_res$iterations
```


Conduct newtonraphson algorithm.
```{r}
newton = function(f, df, x0, tol = 1e-12, maxit = 1000) {
  xs = numeric(maxit + 1)
  xs[1] = x0
  x = x0
  conv_iter = NA_real_
  for (i in 1:maxit) {
    dfx = df(x)
    if (abs(dfx) < .Machine$double.eps) {
      x_new = x
    } else {
      x_new = x - f(x) / dfx
    }
    xs[i + 1] = x_new
    if (is.na(conv_iter) && abs(x_new - x) < tol) conv_iter = i
    x = x_new
  }
  if (is.na(conv_iter)) conv_iter = maxit
  list(root = xs[conv_iter + 1], iterations = conv_iter, values = xs[1:(conv_iter+1)])
}

f = function(x) cos(x) - x
df = function(x) -sin(x) - 1
nr_res = newton(f, df, x0 = 0, tol = 1e-12, maxit = 1000)

signif(nr_res$root, 12)
nr_res$iterations
```

From conducting the code, we see that the newtonraphson method is faster.


## Question 2:

Write a program, using both newtonraphson.r and fixedpoint.r for guidance, 
to implement the secant root-finding method:

$$
x_{n+1} = x_n - f(x_n) \, \frac{x_n - x_{n-1}}{f(x_n) - f(x_{n-1})}.
$$

First test your program by finding the root of the function cos x - x. Next
see how the secant method performs in finding the root of log x - exp(-x)
using $x_0$ = 1 and $x_1$ = 2. Compare its performance with that of the other
two methods. Write a function secant_show.r that plots the sequence of iterates
produced by the secant algorithm.


Create the secant function.
```{r}
secant = function(f, x0, x1, tol = 1e-8, max_iter = 100) {
  iterates = numeric()
  iterates[1] = x0
  iterates[2] = x1
  
  for (i in 1:max_iter) {
    f0 = f(x0)
    f1 = f(x1)
    if (abs(f1 - f0) < .Machine$double.eps) {
      warning("Division by zero encountered.")
      break
    }
    x2 = x1 - f1 * (x1 - x0) / (f1 - f0)
    iterates = c(iterates, x2)
    if (abs(x2 - x1) < tol) {
      cat("Converged in", i, "iterations\n")
      return(list(root = x2, iterates = iterates))
    }
    x0 = x1
    x1 = x2
  }
  warning("Secant method did not converge.")
  return(list(root = NA, iterates = iterates))
}
```


Test 1: f(x) = cos(x) - x
```{r}
f1 = function(x) cos(x) - x
res1 = secant(f1, x0 = 0.5, x1 = 1)
res1$root
```

Test 2: f(x) = log(x) - $e^{-x}$
```{r}
f2 = function(x) log(x) - exp(-x)
res2 = secant(f2, x0 = 1, x1 = 2)
res2$root
```

Use Newton–Raphson to compare.
```{r}
newton_raphson = function(f, fprime, x0, tol = 1e-8, max_iter = 100) {
  x = x0
  for (i in 1:max_iter) {
    x_new = x - f(x) / fprime(x)
    if (abs(x_new - x) < tol) break
    x = x_new
  }
  x
}

fprime2 = function(x) 1/x + exp(-x)
nr_root = newton_raphson(f2, fprime2, x0 = 1)
nr_root
```

Create the plot of the secant method path.
```{r}
secant_show = function(f, iterates) {
  curve(f(x), from = min(iterates) - 0.5, to = max(iterates) + 0.5,
        col = "blue", lwd = 2,
        xlab = "x", ylab = "f(x)",
        main = "Secant Method Path")
  points(iterates, f(iterates), pch = 19, col = "red")
}
secant_show(f2, res2$iterates)
```



## Question 3:

The mean value theorem states that for a function f which is continuous
on [a, b] and differentiable on (a, b), there exists a point $\theta \in$ (a, b)
such that

$$
f'(\theta) = \frac{f(b) - f(a)}{b - a}.
$$

Supposing that f, f', a and b are given, explain how to use a root-finding
algorithm to find $\theta$.


We can rewrite this condition as finding a zero of the function.
$$
g(x) = f'(x) - \frac{f(b) - f(a)}{b - a}
$$

That is, we want $\theta$ as.
$$
g(\theta) = 0.
$$

So, the MVT guarantees that at least one root $\theta \in (a, b)$.
To find $\theta$ numerically, we can use any root-finding methods such as, 
Bisection method, Newton-Raphson method, or Secant method. These methods all 
aim to find x so that g(x) = 0.




## Question 4:

Make your own R function to implement the secant algorithm. Your R function
should have a form:
mysecant <- function(f, x0, x1, tol = 1e-8){
expressions...
}

Test your code by solving logx - $e^{-x}$ + 1 = 0 with respect to x. In doing 
so, you should draw a graph of f(x) = logx - $e^{-x}$ + 1 = 0 over x $\in$ 
[0, 1] and set up two initial values appropriately.

Create the secant function.
```{r}
mysecant = function(f, x0, x1, tol = 1e-8, max_iter = 100) {
  for (i in 1:max_iter) {
    f0 = f(x0)
    f1 = f(x1)
    x2 = x1 - f1 * (x1 - x0) / (f1 - f0)
    if (abs(x2 - x1) < tol) {
      cat("Converged in", i, "iterations\n")
      return(x2)
    }
    x0 = x1
    x1 = x2
  }
  warning("Does not converge")
  return(NA)
}
```

Define the function.
```{r}
f = function(x) log(x) - exp(-x) + 1
```

Plot the function.
```{r}
curve(f(x), from = 0.01, to = 1, col = "blue", lwd = 2,
      ylab = "f(x)", xlab = "x", main = "f(x)")
```

Use the secant method function.
```{r}
root = mysecant(f, x0 = 0.2, x1 = 0.5)
root
f(root)
```




## Question 5:

Let p be the quadratic 
$$
p(x) = c_0 + c_1 x + c_2 x^2.
$$
Simpson’s rule uses a quadratic to approximate a given function f over
two adjacent intervals, then uses the integral of the quadratic to
approximate the integral of the function.

(a) Show that

$$
\int_{-h}^{h} p(x) \, dx = 2h c_0 + \frac{2}{3} c_2 h^3.
$$

We have the quadratic function  

$$
p(x) = c_0 + c_1x + c_2x^2
$$

We want to compute  

$$
\int_{-h}^{h} p(x) \, dx = \int_{-h}^{h} (c_0 + c_1x + c_2x^2) \, dx
$$

Start with integrating by term. 
$$
\int_{-h}^{h} p(x) \, dx 
= c_0 \int_{-h}^{h} dx 
+ c_1 \int_{-h}^{h} x \, dx 
+ c_2 \int_{-h}^{h} x^2 \, dx
$$

And we observe that.
$$
\int_{-h}^{h} dx = 2h, \quad 
\int_{-h}^{h} x \, dx = 0, \quad 
\int_{-h}^{h} x^2 \, dx = \frac{2h^3}{3}
$$

Then substitute the results.
$$
\int_{-h}^{h} p(x) \, dx = 2h c_0 + \frac{2}{3} c_2 h^3
$$



(b) Write down three equations that constrain the quadratic to pass through
the points $(-h, f(-h))$, $(0, f(0))$, and $(h, f(h))$,
then solve them for $c_0$ and $c_2$.

The quadratic.
$$
p(x) = c_0 + c_1 x + c_2 x^2
$$

is required to pass through the three points
$$
(-h,\, f(-h)),\qquad (0,\, f(0)),\qquad (h,\, f(h)).
$$

These give the three equations.
$$
\begin{aligned}
f(-h) &= c_0 - c_1 h + c_2 h^2,\\[4pt]
f(0)  &= c_0\\[4pt]
f(h)  &= c_0 + c_1 h + c_2 h^2
\end{aligned}
$$

From the middle equation we immediately have.
$$
{c_0 = f(0)}
$$

Add the first and third equations to eliminate $c_1$:
$$
f(h) + f(-h) = 2c_0 + 2c_2 h^2.
$$

Substitute $c_0=f(0)$ and solve for $c_2$:
$$
\begin{aligned}
2c_2 h^2 &= f(h) + f(-h) - 2f(0) \\[6pt]
\Rightarrow\quad
c_2 &= \frac{f(h) + f(-h) - 2f(0)}{2h^2}
\end{aligned}
$$

Subtract the first equation from the third to solve for $c_1$:
$$
f(h) - f(-h) = 2 c_1 h
\quad\Longrightarrow\quad
{c_1 = \dfrac{f(h) - f(-h)}{2h}}
$$

So the coefficients determined by interpolation at $-h,0,h$ are
$$
{\,c_0 = f(0),\qquad
c_1 = \dfrac{f(h)-f(-h)}{2h},\qquad
c_2 = \dfrac{f(h)+f(-h)-2f(0)}{2h^2}\,}
$$



(c) Show that

$$
\int_{-h}^{h} p(x) \, dx = \frac{h}{3} \left[ f(-h) + 4f(0) + f(h) \right].
$$

We start from the result of part (a):
$$
\int_{-h}^{h} p(x)\,dx = 2h c_0 + \frac{2}{3} c_2 h^3.
$$

From part (b).
$$
c_0 = f(0), \qquad
c_2 = \frac{f(h)+f(-h)-2f(0)}{2h^2}.
$$

Substitute these into the integral:
$$
\begin{aligned}
\int_{-h}^{h} p(x)\,dx
&= 2h f(0) + \frac{2}{3} h^3 \cdot \frac{f(h)+f(-h)-2f(0)}{2h^2} \\[6pt]
&= 2h f(0) + \frac{h}{3}\bigl(f(h)+f(-h)-2f(0)\bigr) \\[6pt]
&= \frac{h}{3}\bigl(6f(0) + f(h)+f(-h)-2f(0)\bigr) \\[6pt]
&= \frac{h}{3}\bigl(f(-h) + 4f(0) + f(h)\bigr).
\end{aligned}
$$

Then we get.
$$
{\displaystyle
\int_{-h}^{h} p(x)\,dx \;=\; \frac{h}{3}\bigl[f(-h)+4f(0)+f(h)\bigr]
}
$$



## Question 6:

The Riemann approximation to an integral $I = \int_a^b f(x) \, dx$ uses 
rectangles.  
For a partition of size $n$ we have the approximation  

$$
\hat{I} = \sum_{i=0}^{n-1} f(a + ih)h
$$  

where $h = (b - a)/n$ is the width of each interval in the partition.  

What is the difference between the Riemann and trapezoidal approximations 
(using the same partition)?  

It is known that if $-\infty < a < b < \infty$ and $f$ is continuous on 
$[a, b]$, then $\hat{I}$ converges as $n \to \infty$. Using this result, 
conclude that the trapezoidal approximation converges as $n \to \infty$, 
to the same value.

Define simple function to test.
```{r}
f = function(x) exp(-x^2)
a = 0
b = pi
I_true = 2
```

Compute Riemann.
```{r}
riemann = function(f, a, b, n) {
  h = (b - a) / n
  x = seq(a, b - h, by = h)
  sum(f(x)) * h
}
```

Compute trapezoidal.
```{r}
trapezoid = function(f, a, b, n) {
  h = (b - a) / n
  x = seq(a, b, by = h)
  (h/2) * (f(a) + 2 * sum(f(x[-c(1, length(x))])) + f(b))
}
```

Test different partition sizes.
```{r}
n_values = c(10, 50, 100, 1000, 10000)
results = data.frame(
  n = n_values,
  Riemann = sapply(n_values, function(n) riemann(f, a, b, n)),
  Trapezoid = sapply(n_values, function(n) trapezoid(f, a, b, n))
)
```

Compute difference and print results for different partition values.
```{r}
results$Difference = results$Trapezoid - results$Riemann
results
```



## Question 7:

Use the golden section search algorithm to find all of the local maxima of
the function

$$
f(x) =
\begin{cases}
0, & x = 0 \\
|x| \log\left(\frac{|x|}{2}\right) e^{-|x|}, & \text{o/w}
\end{cases}
$$
within the interval [-10, 10].
Hint: plotting the function first will give you a good idea where to look.


Define the function.
```{r}
f = function(x) {
  ifelse(x == 0, 0, abs(x) * log(abs(x) / 2) * exp(-abs(x)))
}
```

Plot function to know where to look.
```{r}
x_vals = seq(-10, 10, length.out = 2000)
y_vals = f(x_vals)
plot(x_vals, y_vals, type = "l", lwd = 2, col = "blue",
     main = "Plot of f(x)",
     xlab = "x", ylab = "f(x)")
```

create a function that uses golden section search algorithm for maxima.
```{r}
golden_search_max = function(f, a, b, tol = 1e-8, maxit = 1000) {
  gr = (sqrt(5) - 1) / 2
  c = b - gr * (b - a)
  d = a + gr * (b - a)
  fc = f(c); fd = f(d)
  iter = 0
  
  while (abs(b - a) > tol && iter < maxit) {
    if (fc < fd) {
      a = c
      c = d
      d = a + gr * (b - a)
      fc = fd
      fd = f(d)
    } else {
      b = d
      d = c
      c = b - gr * (b - a)
      fd = fc
      fc = f(c)
    }
    iter = iter + 1
  }
  x_opt = (a + b) / 2
  list(x = x_opt, f = f(x_opt), iter = iter)
}
```


Looking at the plot, we are going to search for three local maxima.
```{r}
max1 = golden_search_max(f, -10, -1) 
max2 = golden_search_max(f, -1, 1)
max3 = golden_search_max(f, 1, 10) 
```

Print the results.
```{r}
results = data.frame(
  Interval = c("[-10, -1]", "[-1, 1]", "[1, 10]"),
  x_max = c(max1$x, max2$x, max3$x),
  fx_max = c(max1$f, max2$f, max3$f)
)

results
```



