---
title: "Misa Kasparcova STAT 490 HW2"
author: "Misa Kasparcova"
date: "2025-10-04"
output:
  pdf_document: default
  html_document: default
---

Load necessary libraries.
```{r}
library(meta)
library(metafor)
library(dplyr)
library(tidyr)
```

Input data.
```{r}
outcome = c("Duration", "Duration", "Duration", "Duration", "Duration",
            "Duration", "LOS", "LOS", "LOS", "LOS", "LOS", "LOS", "LOS",
            "Analgesic", "Analgesic", "Analgesic", "Analgesic",
            "Analgesic")

study = c(
  "Ahmet Tefekli 2007", "B. Lojanapiwat 2010", "Hemendra N. Shah 2008",
  "Hemendra Shah 2009", "J. Jun-Ou 2010", "Michael Choi 2006",
  "Ahmet Tefekli 2007", "B. Lojanapiwat 2010", "Hemendra N. Shah 2008",
  "Hemendra Shah 2009", "J. Jun-Ou 2010", "Madhu S. Agrawal 2008",
  "Michael Choi 2006", "B. Lojanapiwat 2010", "Hemendra N. Shah 2008",
  "Hemendra Shah 2009", "J. Jun-Ou 2010", "Madhu S. Agrawal 2008")

mean_E = c(60, 49, 51, 52, 47, 82, 38, 85, 35, 34, 82, 22, 37, 39, 150, 103, 
           37, 82)
sd_E = c(9, 24, 10, 23, 17, 18, 10, 23, 11, 17, 24, 4, 24, 35, 97, 116, 31, 24)
n_E = c(17, 45, 33, 454, 43, 12, 17, 45, 33, 454, 43, 101, 12, 45, 33, 454, 
        43, 101)

mean_C = c(76, 57, 47, 68, 59, 73, 67, 129, 44, 56, 106, 54, 38, 75, 246, 250, 
           70, 126)
sd_C = c(10, 20, 16, 34, 18, 15, 22, 54, 22, 62, 35, 5, 24, 32, 167, 132, 36, 
         33)
n_C = c(18, 59, 32, 386, 52, 12, 18, 59, 32, 386, 52, 101, 12, 59, 32, 386, 
        52, 101)

pcnl_data = data.frame(outcome, study, mean_E, sd_E, n_E, mean_C, sd_C, n_C)
pcnl_data
```



## Duration

## Question (a)

Conduct fixed-effect analysis for duration.
```{r}
duration = pcnl_data[pcnl_data$outcome == "Duration",]

meta_fixed_duration = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = duration,
  sm = "MD",        # Mean difference
  common = TRUE,  #Fixed effects
  random = FALSE,
  level = 0.90)

summary(meta_fixed_duration)
```


Conduct random-effect analysis for duration.
```{r}
meta_random_duration = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = duration,
  sm = "MD",             # Mean difference
  common = FALSE,       
  random = TRUE,      # Random effects
  level = 0.90)

summary(meta_random_duration)
```

Most of the studies show statistically significant results at the 90% 
confidence level for a two-tailed test. In contrast, the studies by Hemendra N. 
Shah (2008) and Michael Choi (2006) are not significant, as their confidence 
intervals include zero. Overall, the mean differences across all studies are 
negative, indicating that the effect group’s mean is lower than that of the 
control group.



## Question (b)

Conduct random-effect analysis using DerSimonian-Laird, Paule-Mandel
and restricted MLE methods.


DerSimonian-Laird method for duration.
```{r}
meta_DL_duration = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = duration,
  sm = "MD",                
  common = FALSE,        
  random = TRUE,  # Random effects
  method.tau = "DL",  # DerSimonian-Laird 
  level = 0.90)

summary(meta_DL_duration)
```


Paule-Mandel method for duration.
```{r}
meta_PM_duration = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = duration,
  sm = "MD",
  common = FALSE,
  random = TRUE, # Random effects
  method.tau = "PM", # Paule-Mandel       
  level = 0.90)

summary(meta_PM_duration)
```


Restricted MLE method for duration.
```{r}
meta_REML_duration = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = duration,
  sm = "MD",
  common = FALSE,
  random = TRUE, # Random effects
  method.tau = "REML",  #Restricted MLE
  level = 0.90)

summary(meta_REML_duration)
```

View Tau-squared values.
```{r}
tau_values = data.frame(
  Method = c("DerSimonian-Laird", "Paule-Mandel", "REML"),
  tau_squared   = c(meta_DL_duration$tau^2, meta_PM_duration$tau^2, 
                    meta_REML_duration$tau^2))
print(tau_values)
```

We can observe from the table above that the $\tau^2$ values are all relatively 
similar. The associated p-values are all significant at $\alpha$ = .1, but not 
significant at $\alpha$ = .05.



## Question (c)

Hypotheses:

Null hypothesis ($H_0$): 
$$
\theta_1 = \theta_2 = \dots = \theta_k = \theta
$$

Alternative hypothesis ($H_1$): 
$$
\theta_i, \theta_j \exists\, i, j \text{ such that } \theta_i \ne \theta_j,\ i \ne j
$$

```{r}
Q_value  = meta_fixed_duration$Q
Q_pvalue = meta_fixed_duration$pval.Q
cat("Cochran’s Q =", Q_value, "  p-value =", Q_pvalue, "\n")
```

Observing the p-value that is much smaller than $\alpha$ = 0.1, we can reject 
the null hypothesis and conclude that the random-effects model should be used. 



## LOS

## Question (a)

Conduct fixed-effect analysis for LOS.
```{r}
los = pcnl_data[pcnl_data$outcome == "LOS",]

meta_fixed_los = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = los,
  sm = "MD",        # Mean difference
  common = TRUE,  #Fixed effects
  random = FALSE,
  level = 0.90)

summary(meta_fixed_los)
```


Conduct random-effect analysis for LOS.
```{r}
meta_random_los = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = los,
  sm = "MD",             # Mean difference
  common = FALSE,       
  random = TRUE,      # Random effects
  level = 0.90)

summary(meta_random_los)
```

All of the studies yield significant results for a two-tailed test at 90% 
except for Michael Choi 2006. Additionally, the mean difference is negative,
meaning the mean of the effect group is less than the control group.



## Question (b)

Conduct random-effect analysis using DerSimonian-Laird, Paule-Mandel
and restricted MLE methods.


DerSimonian-Laird method for LOS.
```{r}
meta_DL_los = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = los,
  sm = "MD",                
  common = FALSE,        
  random = TRUE,  # Random effects
  method.tau = "DL",  # DerSimonian-Laird 
  level = 0.90)

summary(meta_DL_los)
```


Paule-Mandel method for LOS.
```{r}
meta_PM_los = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = los,
  sm = "MD",
  common = FALSE,
  random = TRUE, # Random effects
  method.tau = "PM", # Paule-Mandel       
  level = 0.90)

summary(meta_PM_los)
```


Restricted MLE method for LOS.
```{r}
meta_REML_los = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = los,
  sm = "MD",
  common = FALSE,
  random = TRUE, # Random effects
  method.tau = "REML",  #Restricted MLE
  level = 0.90)

summary(meta_REML_los)
```


View Tau-squared values.
```{r}
tau_values = data.frame(
  Method = c("DerSimonian-Laird", "Paule-Mandel", "REML"),
  tau_squared   = c(meta_DL_los$tau^2, meta_PM_los$tau^2, meta_REML_los$tau^2))
print(tau_values)
```

We can observe from the table above that the DerSimonian-Laird $\tau^2$ is less 
than the Paule-Mandel and the REML. The associated p-values are all significant 
at $\alpha$ = .01.


## Question (c)

Hypotheses:

Null hypothesis ($H_0$): 
$$
\theta_1 = \theta_2 = \dots = \theta_k = \theta
$$

Alternative hypothesis ($H_1$): 
$$
\theta_i, \theta_j \exists\, i, j \text{ such that } \theta_i \ne \theta_j,\ i \ne j
$$

```{r}
Q_value  = meta_fixed_los$Q
Q_pvalue = meta_fixed_los$pval.Q
cat("Cochran’s Q =", Q_value, "  p-value =", Q_pvalue, "\n")
```

Observing the p-value that is much smaller than $\alpha$ = 0.1, we can reject 
the null hypothesis and conclude that the random-effects model should be used. 



## Analgesic

## Question (a)

Conduct fixed-effect analysis for Analgesic.
```{r}
ana = pcnl_data[pcnl_data$outcome == "Analgesic",]

meta_fixed_ana = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = ana,
  sm = "MD",        # Mean difference
  common = TRUE,  #Fixed effects
  random = FALSE,
  level = 0.90)

summary(meta_fixed_ana)
```


Conduct random-effect analysis for Analgesic.
```{r}
meta_random_ana = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = ana,
  sm = "MD",             # Mean difference
  common = FALSE,       
  random = TRUE,      # Random effects
  level = 0.90)

summary(meta_random_ana)
```

All of the studies yield significant results for a two-tailed test at 90%. 
Additionally, the mean difference is negative, meaning the mean of the effect
group is less than the control group.



## Question (b)

Conduct random-effect analysis using DerSimonian-Laird, Paule-Mandel
and restricted MLE methods.


DerSimonian-Laird method for Analgesic.
```{r}
meta_DL_ana = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = ana,
  sm = "MD",                
  common = FALSE,        
  random = TRUE,  # Random effects
  method.tau = "DL",  # DerSimonian-Laird 
  level = 0.90)

summary(meta_DL_ana)
```


Paule-Mandel method for Analgesic.
```{r}
meta_PM_ana = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = ana,
  sm = "MD",
  common = FALSE,
  random = TRUE, # Random effects
  method.tau = "PM", # Paule-Mandel       
  level = 0.90)

summary(meta_PM_ana)
```


Restricted MLE method for Analgesic.
```{r}
meta_REML_ana = metacont(
  n.e = n_E, 
  mean.e = mean_E, 
  sd.e = sd_E,
  n.c = n_C, 
  mean.c = mean_C, 
  sd.c = sd_C,
  studlab = study,
  data = ana,
  sm = "MD",
  common = FALSE,
  random = TRUE, # Random effects
  method.tau = "REML",  #Restricted MLE
  level = 0.90)

summary(meta_REML_ana)
```


View Tau-squared values.
```{r}
tau_values = data.frame(
  Method = c("DerSimonian-Laird", "Paule-Mandel", "REML"),
  tau_squared   = c(meta_DL_ana$tau^2, meta_PM_ana$tau^2, meta_REML_ana$tau^2))
print(tau_values)
```

We can observe from the table above that the DerSimonian-Laird $\tau^2$ is 
slightly less than the Paule-Mandel and the REML. The associated p-values 
are all significant at $\alpha$ = .1 and significant at $\alpha$ = .05. 
DerSimonian-Laird is also significant at $\alpha$ = .01.


## Question (c)

Hypotheses:

Null hypothesis ($H_0$): 
$$
\theta_1 = \theta_2 = \dots = \theta_k = \theta
$$

Alternative hypothesis ($H_1$): 
$$
\theta_i, \theta_j \exists\, i, j \text{ such that } \theta_i \ne \theta_j,\ i \ne j
$$



```{r}
Q_value  = meta_fixed_ana$Q
Q_pvalue = meta_fixed_ana$pval.Q
cat("Cochran’s Q =", Q_value, "  p-value =", Q_pvalue, "\n")
```

Observing the p-value that is much smaller than $\alpha$ = 0.1, we can reject 
the null hypothesis and conclude that the random-effects model should be used. 





## Question (d)

Create function for forest plot.
```{r}
analyze_category = function(out, measure = "MD") {
  x = pcnl_data |> filter(outcome == out)
  es = escalc(measure = measure,
               m1i = x$mean_E,
               m2i = x$mean_C,
               sd1i = x$sd_E,
               sd2i = x$sd_C,
               n1i = x$n_E,
               n2i = x$n_C)

  fit = rma.uni(yi, vi, data = es, method = "DL", level = .90)

  forest(fit, 
         slab = x$study,
         xlab = sprintf("Mean Difference (Category %s, DL)",out))
}

par(mfrow = c(2,2))
op = par(mfrow = c(2,2))
on.exit(par(op), add = TRUE)
res = lapply(c("Duration","LOS","Analgesic"), analyze_category, measure = "MD")
par(mfrow = c(1,1))
```


## Question (e)

The results indicate that tubeless and standard PCNL differ significantly 
across all outcome variables at the 0.1 significance level. This is supported 
by the random-effects model p-values, all of which are below $\alpha$ = 0.1, 
regardless of the $\tau^2$ estimation method. Furthermore, the forest plots 
show that the confidence intervals for the random-effects estimates of 
duration, LOS, and analgesic outcomes do not include zero, confirming the 
significance of the findings. Finally, the significant Q test results across 
all three outcome variables justify the use of the random-effects model 
instead of the fixed-effect model, an assumption on which the rest of the 
analysis is based.






