---
title: "Misa Kasparcova STAT 433 HW4"
author: "Misa Kasparcova"
date: "2025-10-30"
output:
  pdf_document: default
  html_document: default
---

## Question 1:

Suppose that we want to find a maximizer of the following function

$$
f(x) = \frac{\log(1 + \log(x))}{\log(1 + x)}, \quad x > 1.
$$

(a) Draw a plot of $f(x)$ over the interval (1, 10) using R.

```{r}
f = function(x) {
  log(1 + log(x)) / (log(1 + x))
}

curve(f, from = 1, to = 10, xlab = "x", ylab = "f(x)")
```



(b) We are going to use Newton’s method. Given $x_n$, provide the update to 
$x_n$, i.e., $x_{n+1}$.


To find the maximum, we first locate the critical points of $f(x)$ where the 
first derivative is zero. Let $f'(x)$ and $f''(x)$ denote the first and 
second derivatives of $f(x)$. The Newton’s method for finding a root of an 
equation $g(x) = 0$ is
$$
x_{n+1} = x_n - \frac{g(x_n)}{g'(x_n)}
$$

Here, we are solving for the point where $f'(x) = 0$, where $g(x) = f'(x)$  
Thus, the Newton’s method update becomes
$$
{
x_{n+1} = x_n - \frac{f'(x_n)}{f''(x_n)}, \quad n = 0, 1, 2, \dots
}
$$



(c)Write an R code to implement Newton’s method for this problem. Show the 
result based on your code. The second derivative of $f(x)$ can be complicated. 
For the second derivative, you could create an R function to compute it using 
the following approximation:

$$
f''(x) \approx \frac{f'(x + h) - f'(x)}{h},
$$

for a small h. If you have no problem calculating the second derivative by hand, 
you should proceed with that method.


Use function from part a.
```{r}
f = function(x) {
  log(1 + log(x)) / (log(1 + x))
}
```


Find first derivative.
```{r}
f_prime = function(x) {
  num = (1 / (x * (1 + log(x)))) * log(1 + x) - log(1 + log(x)) / (1 + x)
  deriv = (log(1 + x))^2
  num / deriv
}
```


Use the approximation for second derivative.
```{r}
f_prime2 = function(x, h = 1e-5) {
  (f_prime(x + h) - f_prime(x)) / h
}
```

Utilize Newton's method.
```{r}
newton_method = function(x0, tol = 1e-8, max_iter = 100) {
  x = x0
  for (i in 1:max_iter) {
    fp = f_prime(x)
    fpp = f_prime2(x)
    x_new = x - fp / fpp
    if (abs(x_new - x) < tol) {
      cat(i, "iterations\n")
      return(x_new)
    }
    x = x_new
  }
  return(x)
}

```

Since x > 1, use $x_0$ = 2 as starting point.
```{r}
x_max = newton_method(2)
x_max
f(x_max)
```


## Question 2:

The probability of recovery from a certain disease is 0.15. Nine people have
contracted the disease. What is the probability that at most 2 of them
recover? What is the expected number that will recover?

Create function for binomial probability.
```{r}
n = 9
p = 0.15

prob = function(n, k) {
  factorial(n) / (factorial(k) * factorial(n - k))
}
```

Obtain probability that at most 2 recover.
```{r}
most_2 = 0
for (k in 0:2) {
  prob_k = prob(n, k) * (p^k) * ((1 - p)^(n - k))
  most_2 = most_2 + prob_k
}

most_2
```

Or utilize pbinom function.
```{r}
second_most_2 = pbinom(2, size = n, prob = p)
second_most_2
```

Obtain expected number that will recover.
```{r}
expected_recover = n * p
expected_recover
```

## Question 3:

Calculating binomial probabilities using a recursive function.

Let $X \sim \text{binom}(k,p)$ and let 
$$
f(x, k, p) = \mathbb{P}(X = x) = \binom{k}{x} p^x (1 - p)^{k-x} 
\text{ for } 0 \le x \le k \text{ and } 0 \le p \le 1
$$

$$
f(0, k, p) = (1 - p)^k
$$

$$
f(x, k, p) = \frac{(k - x + 1)p}{x(1 - p)} f(x - 1, k, p), \quad \text{for } x \ge 1
$$

Use this to write a recursive function binom.pmf(x, k, p) that returns
$f(x, k, p)$.
You can check that your function works by comparing it with the built-in
function dbinom(x, k, p)

Create the function.
```{r}
func = function(x, k, p) {
  if (x == 0) {
    return((1 - p)^k)
  } else {
    solution = ((k - x + 1) * p / (x * (1 - p))) * func(x - 1, k, p)
    return(solution)
  }
}
```

Apply values into function.
```{r}
x = 2
k = 9
p = 0.15

cat("Recursive function:", func(x, k, p))
```

Check the function by comparing with dbinom(x, k, p).
```{r}
x = 2
k = 9
p = 0.15

check = dbinom(x, k, p)
cat("dbinom function:", check)
```



## Question 4:

A random variable $U$ has a $U(a, b)$ distribution if 
$P(U \in (u, v)) = (v - u)/(b - a)$ for all $a \le u \le v \le b$.
Show that if $U \sim U(a, b)$ then so is $a + b - U$.

Let $U \sim U(a, b)$. By definition of the uniform distribution, the 
probability that $U$ lies between any two points $u$ and $v$ such that 
$a \le u \le v \le b$ is given by

$$
P(U \in (u, v)) = \frac{v - u}{b - a}
$$

Define a new random variable
$$
V = a + b - U
$$

We want to show that $V$ also follows a uniform distribution on the same 
interval $[a, b]$; that is, $V \sim U(a, b)$. Since $U$ can take any value 
between $a$ and $b$, we can compute the possible range of $V$:

If $U = a$, then $V = a + b - a = b$.
If $U = b$, then $V = a + b - b = a$.

As $U$ increases from $a$ to $b$, the variable $V = a + b - U$ 
decreases from $b$ to $a$. $V$ also takes values in the interval $[a, b]$.

For any $a \le u \le v \le b$, we have
$$
\begin{aligned}
P(V \in (u, v)) 
&= P(a + b - U \in (u, v)) \\
&= P(U \in (a + b - v, \; a + b - u))
\end{aligned}
$$


Because $U$ is uniformly distributed over $[a, b]$, its probability over any 
subinterval depends only on the length of the interval:

$$
P(U \in (a + b - v, \; a + b - u)) 
= \frac{(a + b - u) - (a + b - v)}{b - a}
= \frac{v - u}{b - a}
$$


Since we have shown that
$$
P(V \in (u, v)) = \frac{v - u}{b - a}
$$

for all $a \le u \le v \le b$, it follows that $V$ has the same probability 
as a $U(a, b)$ random variable.

$$
{V = a + b - U \sim U(a, b)}
$$



## Question 5:

Suppose $X \sim U(0, 1)$ and $Y = X^2$.
Use the cdf of X to show that $P(Y \le y) = \sqrt{y}$ for $0 < y < 1$, and thus
obtain the pdf of $Y$. Hence or otherwise evaluate E($Y$) and Var($Y$).

Let $X \sim U(0,1)$ and define $Y = X^2$. We will find the CDF and pdf of $Y$, then compute $E(Y)$ and $Var(Y)$.


cdf of $Y$ for $0 < y < 1$,
$$
\begin{aligned}
F_Y(y) &= P(Y \le y) = P(X^2 \le y).
\end{aligned}
$$

Since $X$ takes values in [0,1] and is nonnegative, $X^2 \le y$ is equivalent to $X \le \sqrt{y}$
$$
\begin{aligned}
F_Y(y) &= P(X \le \sqrt{y}) = F_X(\sqrt{y}).
\end{aligned}
$$

For $X\sim U(0,1)$ we have $F_X(x)=x$ for $0\le x\le1$. Hence for $0<y<1$
$$
{\,F_Y(y)=\sqrt{y}\, }
$$

To find the pdf of $Y$, differentiate the cdf on $(0,1)$
$$
\begin{aligned}
f_Y(y) &= \frac{d}{dy}F_Y(y)=\frac{d}{dy}\sqrt{y}=\frac{1}{2\sqrt{y}},\qquad 0<y<1
\end{aligned}
$$

pdf is
$$
{\,f_Y(y)=\frac{1}{2\sqrt{y}}\,,\quad 0<y<1\,}
$$


Expectation $E(Y)$ using the pdf of $Y$
$$
\begin{aligned}
E(Y) &= \int_0^1 y\,f_Y(y)\,dy
= \int_0^1 y\cdot\frac{1}{2\sqrt{y}}\,dy
= \frac{1}{2}\int_0^1 \sqrt{y}\,dy.
\end{aligned}
$$

Compute the integral
$$
\frac{1}{2}\int_0^1 y^{1/2}\,dy
= \frac{1}{2}\cdot\frac{2}{3}=\frac{1}{3}.
$$

We get $E(Y)$ as
$$
{\,E(Y)=\frac{1}{3}\,}
$$



Variance $Var(Y)$, compute $E(Y^2)$
$$
\begin{aligned}
E(Y^2) &= \int_0^1 y^2 f_Y(y)\,dy
= \frac{1}{2}\int_0^1 y^{3/2}\,dy
= \frac{1}{2}\cdot\frac{2}{5}=\frac{1}{5}
\end{aligned}
$$

Compute variance
$$
\begin{aligned}
\operatorname{Var}(Y) &= E(Y^2)-[E(Y)]^2
= \frac{1}{5}-\left(\frac{1}{3}\right)^2
= \frac{1}{5}-\frac{1}{9}
= \frac{9-5}{45}=\frac{4}{45}
\end{aligned}
$$




## Question 6:


Telephone calls arrive at a switchboard in accordance with a Poisson process
of rate $\lambda = 5$ per hour.

(a). What is the distribution of $N_1$ = the number of calls that arrive in any
one hour period?

Let telephone calls arrive according to a Poisson process with rate 
$$
\lambda = 5 \text{ calls per hour}
$$

By the definition of a Poisson process, the number of arrivals in a time interval of length $t$ has a Poisson distribution with parameter $\lambda t$
$$
N_1 \sim \text{Poisson}(\lambda t)
$$

For a one-hour period, $t = 1$
$$
N_1 \sim \text{Poisson}(\lambda \cdot 1) = \text{Poisson}(5)
$$

The pmf of $N_1$ is
$$
P(N_1 = k) = \frac{e^{-5} \, 5^k}{k!}, \quad k = 0, 1, 2, \dots
$$


(b). What is the distribution of $N_2$ = the number of calls that arrive in any
half hour period?

Given that the rate is $\lambda = 5$ calls per hour, the expected number of calls in a 
half-hour period is
$$
\lambda t = 5 \times \frac{1}{2} = 2.5.
$$

Using Poisson process, the number of arrivals in a time interval of length $t$ has a Poisson distribution with parameter $\lambda t$
$$
N_2 \sim \text{Poisson}(\lambda t)
$$
The pmf of $N_2$ is
$$
P(N_2 = k) = \frac{e^{-2.5} \, 2.5^k}{k!}, \quad k = 0, 1, 2, \dots
$$


(c). Find the probability that the operator is idle for the next half hour.

The pmf of $N_2$ is
$$
N_2 \sim \text{Poisson}(\lambda t) = \text{Poisson}(5 \times 0.5) = \text{Poisson}(2.5).
$$
The pmf of $N_2$
$$
P(N_2 = 0) = e^{-2.5} \frac{(2.5)^0}{0!} = e^{-2.5}
$$

Compute using code
```{r}
lambda = 5
t = 0.5
lambda_t = lambda * t
p_idle = dpois(0, lambda_t)
p_idle
```




