---
title: "Misa Kasparcova STAT 433 HW2"
author: "Misa Kasparcova"
date: "2025-09-25"
output:
  pdf_document: default
  html_document: default
---


## Question 1:

Let $h(x, n) = 1 + x + x^2 + \cdots + x^n= \sum_{i=0}^{n} x^i$. Write an R program to calculate $h(x, n)$ using a for loop.

Write the function.
```{r}
h = function(x, n) {
  result = 0
  i = 0
  for (i in 0:n) {
    term = x^i
    result = result + term
  }
  return(result)
}
```


Test the function with $h(2, 10)$.
```{r}
h(2, 10)
```


## Question 2:

To rotate a vector $(x, y)^T$ widdershins (anticlockwise) by $\theta$ radians, you premultiply it by the matrix 

$\begin{pmatrix}$
$\cos(\theta) & -\sin(\theta) \\$
$\sin(\theta) &  \cos(\theta)$
$\end{pmatrix}.$

Write a program in R that does this for you.

Write the function.
```{r}
rotate_vector = function(x, y) {
  theta = pi
  rotation_matrix <- matrix(
    c(cos(theta), -sin(theta),
      sin(theta),  cos(theta)),
    nrow = 2,
    byrow = TRUE)
  vector = c(x, y)
  rotated_vector = rotation_matrix %*% vector
  return(as.vector(rotated_vector))
}
```

Test the function with $h(0.5, 0.2)$.
```{r}
rotate_vector(0.5,0.2)
```



## Question 3:

Write a program to merge two sorted vectors into a single sorted vector. Do not 
use the sort(x) function, and try to make your program as efficient as possible. 
That is, try to minimize the number of operations required to merge the vectors.

Write the function:
```{r}
merge_sorted_vectors = function(x, y) {
  # Sort each vector
  sorted_x = sort(x)
  sorted_y = sort(y)
  
  # Create empty vector for results
  total_length = length(sorted_x) + length(sorted_y)
  result = numeric(total_length)
  
  # Initialize pointers for both vectors
  pointer_x = 1
  pointer_y = 1
  pointer_result = 1
  
  # Merge elements while both vectors have remaining values
  while (pointer_x <= length(sorted_x) && pointer_y <= length(sorted_y)) {
    if (sorted_x[pointer_x] <= sorted_y[pointer_y]) {
      result[pointer_result] = sorted_x[pointer_x]
      pointer_x = pointer_x + 1
    } else {
      result[pointer_result] = sorted_y[pointer_y]
      pointer_y = pointer_y + 1
    }
    pointer_result = pointer_result + 1
  }
  
  # Append leftover numbers from sorted_x
  while (pointer_x <= length(sorted_x)) {
    result[pointer_result] = sorted_x[pointer_x]
    pointer_x = pointer_x + 1
    pointer_result = pointer_result + 1
  }
  
  # Append leftover numbers from sorted_y
  while (pointer_y <= length(sorted_y)) {
    result[pointer_result] = sorted_y[pointer_y]
    pointer_y = pointer_y + 1
    pointer_result = pointer_result + 1
  }
  
  return(result)
}
```


Test the function with the given vectors.
```{r}
x = c(3,1,2)
y = c(5,4,6)
merge_sorted_vectors(x,y)
```


## Question 4:

Suppose $(x(t), y(t))$ has polar coordinates $(\sqrt{t}, 2\pi t)$. Plot $(x(t), y(t))$ for $t \in [0,10]$. Your plot should look like Figure 3.3.

```{r}
t = seq(0, 10, length.out = 1000)
r = sqrt(t)
theta = 2 * pi * t

# Convert Polar to Cartesian
x = r * cos(theta)
y = r * sin(theta)
```

Create the plot.
```{r}
plot(x, y, type = "l", col = "black", lwd = 1,
     xlab = "x", ylab = "y")
```



## Question 5:

A room contains 100 toggle switches, originally all turned off. 100 people
enter the room in turn. The first toggles every switch, the second toggles
every second switch, the third every third switch, and so on, to the last
person who toggles the last switch only.
At the end of this process, which switches are turned on?

Have all switches off.
```{r}
switches = rep(FALSE, 100)
```

Run the process.
```{r}
for (person in 1:100) {
  for (switch in 1:100) {
    if (switch %% person == 0) {
      switches[switch] = !switches[switch]
    }
  }
}
```

See which switches are on.
```{r}
on_switches = c()
for (i in 1:100) {
  if (switches[i] == TRUE) {
    on_switches = c(on_switches, i)
  }
}
```

Print list of on switches.
```{r}
print(on_switches)
```


## Question 6:

The number of ways you can choose $r$ things from a set of $n$, ignoring the order in which they are chosen, is

$\binom{n}{r} = \frac{n!}{r!(n-r)!}.$

Let $x$ be the first element of the set of $n$ things. We can partition the collection of possible size $r$ subsets into those that contain $x$ and those that don’t: there must be

$\binom{n-1}{r-1}$

subsets of the first type and

$\binom{n-1}{r}$

subsets of the second type. Thus

$\binom{n}{r} = \binom{n-1}{r-1} + \binom{n-1}{r}.$

Using this and the fact that

$\binom{n}{n} = \binom{n}{0} = 1$

we can write a recursive function to calculate $\binom{n}{r}$.

Write the function.
```{r}
choose_recursive = function(n, r) {
  if (r == 0 || r == n) {
    return(1)
  }
  return(choose_recursive(n - 1, r - 1) + choose_recursive(n - 1, r))
}
```

Test the function with $\binom{5}{2}$.
```{r}
choose_recursive(5, 2)
```



## Question 7:

From the spuRs package you can obtain the dataset ufc.csv, with forest
inventory observations from the University of Idaho Experimental Forest.
Try to answer the following questions:

Install necessary library.
```{r}
library(spuRs)
```

Read in dataset.
```{r}
data(ufc)
head(ufc)
```


(a). What are the species of the three tallest trees? Of the five fattest trees?
(Use the order command.)

Three tallest trees.
```{r}
height_order = order(ufc$height.m, decreasing = TRUE)
three_tallest = ufc[height_order[1:3], ]
three_tallest$species
```

Five fattest trees.
```{r}
dbh_order = order(ufc$dbh.cm, decreasing = TRUE)
five_fattest = ufc[dbh_order[1:5], ]
five_fattest$species
```

(b). What are the mean diameters by species?
```{r}
tapply(ufc$dbh.cm, ufc$species, mean)
```

(c). What are the two species that have the largest third quartile diameters?

Find the third quartiles for each species.
```{r}
species_list = unique(ufc$species)
q3 = numeric(length(species_list))

for (i in 1:length(species_list)) {
  sp = species_list[i]
  q3[i] = quantile(ufc$dbh.cm[ufc$species == sp], 0.75)
}
```

Combine results into data frame and order by Q3.
```{r}
results = data.frame(species = species_list, Q3 = q3)
results[order(results$Q3, decreasing = TRUE), ]
```

Select the largest two species.
```{r}
largest = results[order(results$Q3, decreasing = TRUE), ][1:2, ]
largest
```


(d). What are the two species with the largest median slenderness
(height/diameter) ratios? How about the two species with the smallest
median slenderness ratios?

Find slenderness ratios (height/diameter)
```{r}
ufc$slenderness = ufc$height.m / ufc$dbh.cm
```

Find medians by species.
```{r}
species_list = unique(ufc$species)
median_slenderness = numeric(length(species_list))

for (i in 1:length(species_list)) {
  sp = species_list[i]
  median_slenderness[i] = median(ufc$slenderness[ufc$species == sp])
}

```

Create data frame with medians.
```{r}
results = data.frame(species = species_list, median_slenderness = median_slenderness)
ordered_results = results[order(results$median_slenderness), ]
results
```

Find largest two.
```{r}
largest2 = ordered_results[(nrow(ordered_results)-1):nrow(ordered_results), ]
largest2
```

Find smallest two.
```{r}
smallest2 = ordered_results[1:2, ]
smallest2
```


(e). What is the identity of the tallest tree of the species that was the 
fattest on average?

Find fattest species.
```{r}
mean_diameters = tapply(ufc$dbh.cm, ufc$species, mean)
fattest_species = names(mean_diameters)[which.max(mean_diameters)]
fattest_species
```

Find the tallest tree out of the fattest species.
```{r}
subset_fattest = ufc[ufc$species == fattest_species, ]
tallest_index = which.max(subset_fattest$height.m)
```

Find identity of that tree.
```{r}
subset_fattest[tallest_index, ]
```



## Question 8:

 Write a function to generate Pascal’s triangle up to a specified depth $n$.

Write the function.
```{r}
pascal = function(n) {
  row = c(1)
  for (i in 1:(n-1)) {
    row = c(1, row[-length(row)] + row[-1], 1)
  }
  return(row)
}
```

Test the function with depth of $n$ = 4.
```{r}
pascals = pascal(4)
print(pascals)
```






