---
title: "Misa Kasparcova STAT 433 HW6 R Code"
author: "Misa Kasparcova"
date: "2025-11-23"
output:
  pdf_document: default
  html_document: default
---

For all questions, generate 10000 random numbers from the corresponding 
probability distributions. For each question, write R code, execute it, 
report the results.


## Question 1

Write R code to compute a Monte Carlo estimate $\hat{\theta}$ of

$$
\theta = \int_{0}^{\pi/3} \sin(t)\, dt,
$$

and compare your result with the exact value of the integral.

Set seed:
```{r}
set.seed(123)
```

Generate 10000 random numbers from distribution:
```{r}
n = 10000
u = runif(n, 0, pi/3)
```

Compute Monte Carlo estimate $\hat{\theta}$:
```{r}
mean = mean(sin(u))
theta_hat = (pi/3) * mean
theta_hat
```

Compute exact value of integral:
```{r}
func = function(t) sin(t)
exact_value = integrate(func, lower = 0, upper = pi/3)$value
exact_value
```

The Monte Carlo estimate of the integral of 0.49808 is extremely close to the 
exact value of the integral of 0.5. This shows that the simulation accurately 
approximates the true value sample from 10,000 randomly generated numbers.



## Question 2

Write R code to compute a Monte Carlo estimate $\hat{\theta}$ of

$$
\theta = \int_{0}^{1.5} e^{-x}\, dx,
$$

by using random samples from $(i)$ $U(0, 1.5)$ and $(ii)$ $U(0, 1)$, 
respectively. Provide estimate of the variance of $\hat{\theta}$ for each case.


For $(i)$ $U(0, 1.5)$:

Set seed:
```{r}
set.seed(123)
```

Generate 10000 random numbers from distribution:
```{r}
n = 10000
u2 = runif(n, 0, 1.5)
```

Obtain $\hat{\theta}$ and variance of $\hat{\theta}$:
```{r}
mean = mean(exp(-u2))
var = var(exp(-u2))

theta_hat2 = 1.5 * mean
var_hat2 = (1.5^2) * var / n

theta_hat2
var_hat2
```

Compute exact value of integral:
```{r}
func = function(x) exp(-x)
exact_value = integrate(func, lower = 0, upper = 1.5)$value
exact_value
```

Compute exact variance of integral to check:
```{r}
n = 10000

E1 = (1 - exp(-1.5)) / 1.5
E2 = (1 - exp(-3)) / 3

var_exact_1 = (1.5^2 / n) * (E2 - E1^2)
var_exact_1
```




For $(ii)$ $U(0, 1)$:

Set seed:
```{r}
set.seed(123)
```

Generate 10000 random numbers from distribution:
```{r}
n = 10000
u3 = runif(n, 0, 1)
```

Obtain $\hat{\theta}$ and variance of $\hat{\theta}$:
```{r}
mean3 = mean(exp(-u3))
var3 = var(exp(-u3))

theta_hat3 = 1 * mean3
var_hat3 = (1^2) * var3 / n

theta_hat3
var_hat3
```

Compute exact value of integral to check:
```{r}
func = function(x) exp(-x)
exact_value = integrate(func, lower = 0, upper = 1)$value
exact_value
```

Compute exact variance of integral to check:
```{r}
n = 10000

E1 = (1 - exp(-1)) / 1
E2 = (1 - exp(-2)) / 2

var_exact_2 = (1^2 / n) * (E2 - E1^2)
var_exact_2
```

We observe with both the $U(0, 1.5)$ and $U(0, 1)$ case, the Monte Carlo 
estimate of $\hat{\theta}$ and its variance are very similar to the exact 
value and variance of the integral.



## Question 3

Write R code to estimate the integral

$$
\theta = \int_{0}^{1} \sqrt{\,1 - x^{2}\,}\, dx,
$$

using importance sampling with the following density function:

$$
h(x) = 4x^{3}, \qquad 0 < x < 1.
$$

Set seed:
```{r}
set.seed(123)
```

Generate 10000 random numbers from distribution and obtain inverse of x:
```{r}
n = 10000
u4 = runif(n, 0, 1)
x = u^(1/4)
x = pmin(x, 1)
```


Define $\theta$ and $h(x)$:
```{r}
theta = sqrt(1 - x^2)
hx = 4 * x^3
```


Obtain $\hat{\theta}$ estimate:
```{r}
theta_hat3 = mean(theta / hx)
theta_hat3
```

Compute exact value of integral to check:
```{r}
func = function(x) sqrt(1 - x^2)
exact_value = integrate(func, lower = 0, upper = 1)$value
exact_value
```

The Monte Carlo estimate of the integral of 0.6798 is relatively close to the 
exact value of the integral of 0.7854.



## Question 4

Make R code to estimate the integral

$$
\theta = \int_{0}^{\infty} x^{2}\, \frac{e^{-x^{2}/2}}{\sqrt{2\pi}}\, dx,
$$

using importance sampling where the sampling density is set to $expon(1)$.


Set seed:
```{r}
set.seed(123)
```


Generate 10000 random numbers and sample from expon(1):
```{r}
n = 10000
x = rexp(n, rate = 1)
```


Define $f(x)$ and $h(x)$:
```{r}
fx = x^2 * exp(-x^2 / 2) / sqrt(2 * pi)
hx = dexp(x, rate = 1)
```

Obtain estimate for the integral:
```{r}
w = fx / hx
theta_hat4 = mean(w)
theta_hat4
```

Compute exact value of integral to check:
```{r}
func = function(x) x^2 * exp(-x^2 / 2) / sqrt(2 * pi)
exact_value = integrate(func, lower = 0, upper = Inf)$value
exact_value
```

The Monte Carlo estimate of the integral of 0.4401 is close to the exact value 
of the integral of 0.5.
